---
title: "Lab 10: Comparative Structure Analysis"
author: "Trevor Hoang (A16371830)"
format: html
---

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifier and ends in PCA analysis.

```{r}
library(bio3d)

```

##Step 1. 
Retrieve a sequence for the protein that we are interested in. We will take ADK "1ake_A" 

```{r}
id = "1ake_A"
aa = get.seq(id)
aa
```

##Step. 2 Blast Search

Run a BLAST search of the PDB for all related sequeced to our input `aa`

```{r}
blast = blast.pdb(aa)
```

```{r}
hits = plot(blast)
```

##Step 3. Download all Structures

There are our "top hits" i.e. all structures in the PDB database related to our input sequences.
```{r}
hits$pdb.id
```

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

##Step 4. Align and Superimpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

##Step 5. PCA

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships.

```{r}
pc = pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of he main conformational changes captured by PC1. We will use the `mktrj()` function for this.

```{r}
mktrj(pc, file= "pca_results.pdb")
```

Back of the envelope comparison of the PDB size to UniProt

```{r}
uniprot = 253206171
pdb= 231029
pdb/uniprot *100
```









