---
title: "Lab 11: Alphafold Analysis"
author: "Trevor Hoang (A16371830)"
format: html
---

Here we analyze our AlphaFold structure prediction models. The input directory/folder comes from the ColabFold server.

```{r}
results = "hivpr_dimer_23119"
```

```{r}
pdb_files <- list.files(path=results,
                        pattern="*.pdb",
                        full.names = TRUE)

basename(pdb_files)
```

I will use the Bio3D package for analysis

```{r}
library(bio3d)

pdbs = pdbaln(pdb_files, fit=T, exefile = "msa")
```

Align and superimpose

```{r}
pdbs
```

Using `rmsd()` function to calculate the RMSD between all pairs models
```{r}
rd = rmsd(pdbs,fit=T)

range(rd)
```

Drawing a heatmap

```{r}
library(pheatmap)

colnames(rd) = paste0("m", 1:5)
rownames(rd) = paste0("m", 1:5)
pheatmap(rd)
```

```{r}
pdb = read.pdb("1hsg")
```

```{r}
core = core.find(pdbs)

core.inds = print(core, vol=0.5)
```

```{r}
xyz = pdbfit(pdbs, core.inds, outpath ="corefit_structures")
```

Examining the RMSG between positions of the structure

```{r}
rf = rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```

## Predicted Alignment Error

Prediction for model 1

```{r}
library(jsonlite)

pae_files <- list.files(path=results,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)

pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)

```

```{r}
head(pae1$plddt)
```

```{r}
pae1$max_pae
pae5$max_pae
```

Plotting PAE scores
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

##Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file

```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

Finding the amount of sequences in the alignment
```{r}
dim(aln$ali)

```
Scoring the residue consercation

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```





